Learning to Pivot with Adversarial Networks

### 2017-05-16 tex/nips/nips_2017.tex
minor edits

@@

@@

Assuming a probability model $p(X,Y,Z)$, where $X$ are the data, $Y$ are the
target labels, and $Z$ are the nuisance parameters, we consider the problem of
learning a predictive model $f(X)$ for $Y$ conditional on the observed values of $X$
that is robust to uncertainty in the unknown value of $Z$. More specifically, we
introduce a learning procedure based on adversarial networks
for enforcing that  $f(X)$ is a pivot, i.e., a quantity whose distribution
does not depend on $Z$.
We derive theoretical results proving that the proposed procedure
converges towards a predictive model that is both optimal and statistically independent of the
nuisance parameters (if that model exists) or for which one can tune a
trade-off between power and robustness (e.g., as driven by the requirements of a higher level statistical analysis).
Finally, we demonstrate the
effectiveness of this approach with a toy example and an example from particle physics.

@@

We begin with a family of data generation processes $p(X,Y,Z)$, where $X \in
\mathcal{X}$ are the data, $Y\in \mathcal{Y}$ are the target labels, and $Z\in
\mathcal{Z}$ are the nuisance parameters that can be continuous or categorical.
% We assume training data $\{x_i, y_i, z_i\}_{i=1}^N$.
Let us assume that prior to incorporating the effect of uncertainty
in $Z$, our goal is to learn a regression function $f : \mathcal{X} \mapsto
{\cal S}$ of parameters $\theta_f$ (e.g., a neural network-based probabilistic
classifier) that minimizes a loss ${\cal L}_f(\theta_f)$ (e.g., the
cross-entropy). In classification, values $\smash{s \in {\cal S} = \mathbb{R}^{|{\cal Y}|}}$
correspond to scores used for mapping hard predictions $y \in {\cal Y}$, while ${\cal S} = {\cal Y}$ for regression.

@@

Joint training of adversarial networks was first proposed by \citep{goodfellow2014generative} as a
way to build a generative model capable of producing samples from random noise
$z$. More specifically, the authors pit a generative model $g:
\mathbb{R}^n \mapsto \mathbb{R}^p$ against an adversary classifier $d :
\mathbb{R}^p \mapsto [0, 1]$ whose antagonistic objective is to recognize
real data $X$ from generated data $g(Z)$. Both models $g$ and $d$ are trained
simultaneously, in such a way that $g$ learns to produce samples that are
difficult to identify by $d$, while $d$ incrementally adapts to changes in $g$.
At the equilibrium, $g$ models a distribution whose samples can be identified by
$d$ only by chance. That is, assuming enough capacity in $d$ and  $g$, the
distribution of $g(Z)$ eventually converges towards the real distribution
of $X$.

In this work, we repurpose adversarial networks as a means to constrain the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. As
illustrated in Fig.~\ref{fig:architecture}, we pit $f$ against an adversary
model $r := p_{\theta_r}(z | f(X;\theta_f)=s)$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$. This model takes  as input
realizations $s$ of $f(X; \theta_f)$ and produces as output a function
modeling the posterior probability density $p_{\theta_r}(z | f(X;\theta_f)=s)$.
Intuitively, if $p(f(X; \theta_f)=s|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)=s|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented as a
probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal Z|}}$ whose
$j^\textrm{th}$ output (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
$p_{\theta_r}(z_j|f(X;\theta_f)=s)$. Similarly, if $Z$ takes continuous values,
then we can model the posterior probability density $p(z | f(X;\theta_f)=s)$
with a sufficiently flexible  parametric family of distributions $\mathcal{P}(\gamma_1, \gamma_2, \dots)$,
where the parameters $\gamma_j$ depend on $f(X, \theta_f)$ and $\theta_r$.
The adversary $r$ may
take any form, i.e. it does not need to be a neural network, as long as it exposes a
differentiable function $p_{\theta_r}(z|f(X;\theta_f)=s)$ of sufficient capacity
to represent the true distribution. Fig.~\ref{fig:architecture} illustrates
a concrete example where $p_{\theta_r}(z | f(X;\theta_f)=s)$ is a mixture of
gaussians, as modeled with a mixture density network~\citep{bishop1994mixture}).
The $j^\textrm{th}$ output corresponds to the estimated value of the
corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
variance and mixing coefficients of its components). The estimated probability density
$p_{\theta_r}(z|f(X;\theta_f)=s)$ can then be evaluated for any $z \in {\cal Z}$ and any score $s \in {\cal S}$.

%If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented e.g. as a
%probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal Z|}}$ whose
%output $j$ (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
%$p_{\theta_r}(z_j|f(X;\theta_f)=s)$. Similarly, if $Z$ takes continuous values and
%if we assume a flexible enough parametric distribution ${\cal P}$ for $Z|f(X;\theta_f)=s$
%(e.g., a mixture of gaussians, as modeled with a mixture
%density network~\citep{bishop1994mixture}), then $p_{\theta_r}$ can be
%represented e.g. as a network whose output $j$ is the estimated value of the
%corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
%variance and mixing coefficients of its components). As in
%\citep{nix1994estimating,bishop1994mixture}, the estimated probability density
%$p_{\theta_r}(z|f(X;\theta_f)=s)$ can then be evaluated for any $z \in {\cal Z}$ and any score $s \in {\cal S}$.
%As further explained in the next section, let us note that the adversary $r$ may
%take any form, i.e. it does not need to be a neural network, as long as it exposes a
%differentiable function $p_{\theta_r}(z|f(X;\theta_f)=s)$ of sufficient capacity
%to represent the true distribution.

As with generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_f, \theta_r)
\end{equation}
that we optimize by finding the minimax solution
\begin{equation}\label{eqn:min_thetaf}
    \smash{\hat\theta_f, \hat\theta_r} = \arg \min_{\theta_f} \max_{\theta_r} E(\theta_f, \theta_r).
\end{equation}
Without loss of generality, the adversarial training procedure to obtain
$(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of a binary classifier $f :
\mathbb{R}^p \mapsto [0,1]$ modeling $p(Y=1|X)$. For reasons further explained
in Sec.~\ref{sec:theory}, ${\cal L}_f$ and ${\cal L}_r$  are respectively set to the
expected value of the
negative log-likelihood of $Y|X$ under $f$ and of $Z|f(X;\theta_f)$ under
$r$:
\begin{align}
    {\cal L}_f(\theta_f) &= \mathbb{E}_{x \sim X}  \mathbb{E}_{y \sim Y|x} [ -\log p_{\theta_f} (y|x) ], \\
    {\cal L}_r(\theta_f, \theta_r) &= \mathbb{E}_{s \sim f(X;\theta_f)}  \mathbb{E}_{z \sim Z|s} [-\log p_{\theta_r} (z|s)].
\end{align}
The optimization algorithm consists in using stochastic gradient descent
alternatively for solving Eqn.~\ref{eqn:min_thetaf}.
Finally, in the case of a class conditional pivot, the settings are the
same, except that the adversarial term ${\cal L}_r(\theta_f, \theta_r)$ is restricted to $Y=y$.

@@

As a guiding toy example, let us consider the binary classification of 2D
data drawn from multivariate gaussians with equal priors, such that
\begin{align}
    x &\sim {\cal N}\left ((0,0), \begin{bmatrix}
                              1 & -0.5 \\
                              -0.5 & 1
                            \end{bmatrix}\right) &\text{ when } Y=0, \\
    x &\sim {\cal N}\left ((1,1+Z),  \begin{bmatrix}
                              1 & 0 \\
                              0 & 1
                             \end{bmatrix}\right) &\text{ when } Y=1.
\end{align}
The continuous nuisance parameter $Z$ represents in this case our
uncertainty about the exact location of the mean of the second gaussian. Our goal is to
build a classifier $f(\cdot;\theta_f)$ for predicting $Y$ given $X$, but such that
the probability distribution of $f(X;\theta_f)$ is invariant with respect to the
nuisance parameter $Z$.

@@

In machine learning, learning a pivotal quantity can be related to the problem
of domain adaptation
\citep{blitzer2006domain,pan2011domain,gopalan2011domain,gong2013connecting,baktashmotlagh2013unsupervised,ajakan2014domain,ganin2014unsupervised},
where the goal is often stated as trying to learn a domain-invariant
representation of the data. Likewise, our method also relates to the problem of
enforcing fairness in classification
\citep{zemel2013learning,feldman2015certifying,EdwardsS15,zafar2015fairness},
which is stated as learning a classifier that is independent of some chosen
attribute such as gender, color or age. For both families of methods, the
problem can equivalently be stated as learning a classifier which is a pivotal
quantity with respect to either the domain or the selected feature. As an
example, unsupervised domain adaptation with labeled data from a source domain
and unlabeled data from a target domain can be recast as learning a predictive
model $f$ (i.e., trained to minimize ${\cal L}_f$ evaluated on labeled source
data only) that is also a pivot with respect to the domain $Z$ (i.e., trained to
maximize ${\cal L}_r$ evaluated on both source and target data). In this
context, \citep{ganin2014unsupervised,EdwardsS15} are certainly among the
closest to our work, in which domain invariance and fairness are enforced
through an adversarial minimax setup composed of a classifier and an adversary
discriminator. Following this line of work and without making assumptions on the
prior $p(Z)$, our method can be regarded as a generalization that also supports
a continuously parametrized family of domains or as enforcing fairness over
continuous attributes.

>>>>

\usepackage{color}
\newcommand{\glnote}[1]{\textcolor{red}{[GL: #1]}}

@@

\glnote{improve abstract}

@@

Assuming a probability model $p(X,Y,Z)$, where $X$ are the data, $Y$ are the
target labels, and $Z$ are the nuisance parameters, we consider the problem of
learning a predictive model $f(X)$ for $Y$ conditional on the observed values of $X$
that is robust to uncertainty in the unknown value of $Z$. More specifically, we
introduce a learning procedure based on adversarial networks~\citep{goodfellow2014generative}
for enforcing that  $f(X)$ is a pivot, i.e., a quantity whose distribution
does not depend on $Z$.
We derive theoretical results proving that the proposed procedure
converges towards a predictive model that is both optimal and statistically independent of the
nuisance parameters (if that model exists) or for which one can tune a
trade-off between power and robustness (e.g., as driven by the requirements of a higher level statistical analysis).
Finally, we demonstrate the
effectiveness of this approach with a toy example and an example from particle physics.

@@

We begin with a family of data generation processes $p(X,Y,Z)$, where $X \in
\mathcal{X}$ are the data, $Y\in \mathcal{Y}$ are the target labels, and $Z\in
\mathcal{Z}$ are the nuisance parameters that can be continuous or categorical.
% We assume training data $\{x_i, y_i, z_i\}_{i=1}^N$.
Let us assume that prior to incorporating the effect of uncertainty
in $Z$, our goal is to learn a regression function $f : \mathcal{X} \to
{\cal S}$ with parameters $\theta_f$ (e.g., a neural network-based probabilistic
classifier) that minimizes a loss ${\cal L}_f(\theta_f)$ (e.g., the
cross-entropy). In classification, values $\smash{s \in {\cal S} = \mathbb{R}^{|{\cal Y}|}}$
correspond to scores used for mapping hard predictions $y \in {\cal Y}$, while ${\cal S} = {\cal Y}$ for regression.

@@

Joint training of adversarial networks was first proposed by \citep{goodfellow2014generative} as a
way to build a generative model capable of producing samples from random noise
$z$. More specifically, the authors pit a generative model $g:
\mathbb{R}^n \to \mathbb{R}^p$ against an adversarial classifier $d :
\mathbb{R}^p \to [0, 1]$ whose antagonistic objective is to recognize
real data $X$ from generated data $g(Z)$. Both models $g$ and $d$ are trained
simultaneously, in such a way that $g$ learns to produce samples that are
difficult to identify by $d$, while $d$ incrementally adapts to changes in $g$.
At the equilibrium, $g$ models a distribution whose samples can be identified by
$d$ only by chance. That is, assuming enough capacity in $d$ and  $g$, the
distribution of $g(Z)$ eventually converges towards the real distribution
of $X$.

In this work, we repurpose adversarial networks as a means to constrain the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. As
illustrated in Fig.~\ref{fig:architecture}, we pit $f$ against an adversarial
model $r := p_{\theta_r}(z | f(X;\theta_f)=s)$ with parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$. This model takes  as input
realizations $s$ of $f(X; \theta_f)$ and produces as output a function
modeling the posterior probability density $p_{\theta_r}(z | f(X;\theta_f)=s)$.
Intuitively, if $p(f(X; \theta_f)=s|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)=s|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented as a
probabilistic classifier $\mathbb{R} \to \mathbb{R}^{|{\cal Z|}}$ whose
$j^\textrm{th}$ output (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
$p_{\theta_r}(z_j|f(X;\theta_f)=s)$. Similarly, if $Z$ takes continuous values,
then we can model the posterior probability density $p(z | f(X;\theta_f)=s)$
with a sufficiently flexible  parametric family of distributions $\mathcal{P}(\gamma_1, \gamma_2, \dots)$,
where the parameters $\gamma_j$ depend on $f(X, \theta_f)$ and $\theta_r$.
The adversary $r$ may
take any form, i.e. it does not need to be a neural network, as long as it exposes a
differentiable function $p_{\theta_r}(z|f(X;\theta_f)=s)$ of sufficient capacity
to represent the true distribution. Fig.~\ref{fig:architecture} illustrates
a concrete example where $p_{\theta_r}(z | f(X;\theta_f)=s)$ is a mixture of
gaussians, as modeled with a mixture density network~\citep{bishop1994mixture}).
The $j^\textrm{th}$ output corresponds to the estimated value of the
corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
variance and mixing coefficients of its components). The estimated probability density
$p_{\theta_r}(z|f(X;\theta_f)=s)$ can then be evaluated for any $z \in {\cal Z}$ and any score $s \in {\cal S}$.

%If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented e.g. as a
%probabilistic classifier $\mathbb{R} \to \mathbb{R}^{|{\cal Z|}}$ whose
%output $j$ (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
%$p_{\theta_r}(z_j|f(X;\theta_f)=s)$. Similarly, if $Z$ takes continuous values and
%if we assume a flexible enough parametric distribution ${\cal P}$ for $Z|f(X;\theta_f)=s$
%(e.g., a mixture of gaussians, as modeled with a mixture
%density network~\citep{bishop1994mixture}), then $p_{\theta_r}$ can be
%represented e.g. as a network whose output $j$ is the estimated value of the
%corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
%variance and mixing coefficients of its components). As in
%\citep{nix1994estimating,bishop1994mixture}, the estimated probability density
%$p_{\theta_r}(z|f(X;\theta_f)=s)$ can then be evaluated for any $z \in {\cal Z}$ and any score $s \in {\cal S}$.
%As further explained in the next section, let us note that the adversary $r$ may
%take any form, i.e. it does not need to be a neural network, as long as it exposes a
%differentiable function $p_{\theta_r}(z|f(X;\theta_f)=s)$ of sufficient capacity
%to represent the true distribution.

As with generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_f, \theta_r)
\end{equation}
that we optimize by finding the minimax solution
\begin{equation}\label{eqn:min_thetaf}
    \smash{\hat\theta_f, \hat\theta_r} = \arg \min_{\theta_f} \max_{\theta_r} E(\theta_f, \theta_r).
\end{equation}
Without loss of generality, the adversarial training procedure to obtain
$(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of a binary classifier $f :
\mathbb{R}^p \to [0,1]$ modeling $p(Y=1|X)$. For reasons further explained
in Sec.~\ref{sec:theory}, ${\cal L}_f$ and ${\cal L}_r$  are respectively set to the
expected value of the
negative log-likelihood of $Y|X$ under $f$ and of $Z|f(X;\theta_f)$ under
$r$:
\begin{align}
    {\cal L}_f(\theta_f) &= \mathbb{E}_{x \sim X}  \mathbb{E}_{y \sim Y|x} [ -\log p_{\theta_f} (y|x) ], \\
    {\cal L}_r(\theta_f, \theta_r) &= \mathbb{E}_{s \sim f(X;\theta_f)}  \mathbb{E}_{z \sim Z|s} [-\log p_{\theta_r} (z|s)].
\end{align}
The optimization algorithm consists in using stochastic gradient descent
alternatively for solving Eqn.~\ref{eqn:min_thetaf}.
Finally, in the case of a class conditional pivot, the settings are the
same, except that the adversarial term ${\cal L}_r(\theta_f, \theta_r)$ is restricted to $Y=y$.

@@

As a guiding toy example, let us consider the binary classification of 2D
data drawn from multivariate gaussians with equal priors, such that
\begin{align}
    x &\sim {\cal N}\left ((0,0), \begin{bmatrix}
                              1 & -0.5 \\
                              -0.5 & 1
                            \end{bmatrix}\right) &\text{ when } Y=0, \\
    x|Z=z &\sim {\cal N}\left ((1,1+z),  \begin{bmatrix}
                              1 & 0 \\
                              0 & 1
                             \end{bmatrix}\right) &\text{ when } Y=1.
\end{align}
The continuous nuisance parameter $Z$ represents in this case our
uncertainty about the exact location of the mean of the second gaussian. Our goal is to
build a classifier $f(\cdot;\theta_f)$ for predicting $Y$ given $X$, but such that
the probability distribution of $f(X;\theta_f)$ is invariant with respect to the
nuisance parameter $Z$.

@@

In machine learning, learning a pivotal quantity can be related to the problem
of domain adaptation
\citep{blitzer2006domain,pan2011domain,gopalan2011domain,gong2013connecting,baktashmotlagh2013unsupervised,ajakan2014domain,ganin2014unsupervised},
where the goal is often stated as trying to learn a domain-invariant
representation of the data. Likewise, our method also relates to the problem of
enforcing fairness in classification
\citep{zemel2013learning,feldman2015certifying,EdwardsS15,zafar2015fairness},
which is stated as learning a classifier that is independent of some chosen
attribute such as gender, color or age. For both families of methods, the
problem can equivalently be stated as learning a classifier which is a pivotal
quantity with respect to either the domain or the selected feature. As an
example, unsupervised domain adaptation with labeled data from a source domain
and unlabeled data from a target domain can be recast as learning a predictive
model $f$ (i.e., trained to minimize ${\cal L}_f$ evaluated on labeled source
data only) that is also a pivot with respect to the domain $Z$ (i.e., trained to
maximize ${\cal L}_r$ evaluated on both source and target data). In this
context, \citep{ganin2014unsupervised,EdwardsS15} are certainly among the
closest to our work, in which domain invariance and fairness are enforced
through an adversarial minimax setup composed of a classifier and an adversarial
discriminator. Following this line of work and without making assumptions on the
prior $p(Z)$, our method can be regarded as a generalization that also supports
a continuously parametrized family of domains or as enforcing fairness over
continuous attributes.
